A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-22 text/informal-overview.tex
Rewrite Sections 1 and 2.

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{stage variable}\footnote{%
  In Hanada and Igarash~\cite{Hanada2014}, it was called a \textit{transition variable}, which is derived from correspondence to modal logic, studied by Tsukada and Igarashi~\cite{Tsukada}.} and
a sequence of stage variables is called \textit{stage}.
Intuitively, they represent the depth of nested brackets.
Stage variables can be abstracted by $\Lambda\alpha.M$ and instantiated by
an application $M\ A$ to stages.
For example, $\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10) 5)$ is a code value, where \(\alpha\)
is abstracted.  If it is applied to \(\varepsilon\), which is the empty sequence standing for zero,
\(\TB_\alpha\) becomes \(\TB_\varepsilon\), which stands for zero brackets,
and so \(\TB_\alpha\) disappears.  As a result, the application reduces to (unquoted) \((\lambda x:\I.x+10) 5\) and to 15.  If \(\alpha\) is instantiated with, say, \(\beta\gamma\), then
\(\TB_\alpha\) becomes nested brackets \(\TB_{\beta} \TB_\gamma\).  These are expressed by
the following reduction rule:
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
where stage substitution \([\alpha \mapsto A]\) manipulates nesting of
\(\TB\) and \(\TBL\) (and also \(\%_\alpha\) as we see later).  As we
show above, application to $\varepsilon$ corresponds to \texttt{run}.

Cross-stage persistence (CSP), which is an important feature of \LTP,
is a primitive to embed values (not necessarily code values) into a
code value.
For example,
\begin{verbatim}
   let y = 1 + 2 in
   .< y * 2 >.    (* .< e >. stands for brackets *)
\end{verbatim}
evaluates to \verb|.< 3 * 2 >.|, not \verb|.< plusone 2 * 2 >.|.
Here, CSP is implicitly applied to the occurrence of \texttt{y}
and embeds its value \texttt{3} into the code.
In \LTP, CSP is written explicitly and $\%$ is a symbol for CSP.
For example, the above MetaOCaml expression can be expressed
in \LTP as follows:
\[
  (\lambda x:\I.\Lambda\alpha.(\TB_\alpha ((\%_\alpha y) * 2)))\ (1 + 2)
\]
It evaluates to \(\Lambda\alpha.(\TB_\alpha ((\%_\alpha 3) * 2))\), in
which $\%_\alpha 3$ means that it waits for the surrounding code to be
run (by an application to $\varepsilon$).  If it is applied to
\(\varepsilon\), substitution of \(\varepsilon\) for \(\alpha\)
eliminates \(\TB_\alpha\) and \(\%_\alpha\) and so \(3 * 2\) is
obtained.

The type system of \LTP uses code types---the type of code of type
\(\tau\) is written \(\TW_\alpha \tau\)---for typing \(\TB_\alpha\),
\(\TBL_\alpha\) and \(\%_\alpha\).  It takes stages into account: A
variable declaration in a type environment is associated with its
declared stage as well as its type and a type judgement of \LTP is of
the form $\G \vdash M : \tau @ A$, in which $A$ stands for the stage
of term $M$.  For example,
$y:\I@\alpha \vdash (\lambda x:\I.y) : \I \to \I @ \alpha$ holds, but
$y:\I@\alpha \vdash (\lambda x:\I.y) : \I \to \I @ \varepsilon$ does
not because the latter uses $y$ at stage \(\varepsilon\) but $y$ is
declared at $\alpha$.  Quotation \(\TB_\alpha M\) is given type
\(\TW_\alpha \tau\) at stage \(A\) if \(M\) is given type \(\tau\) at
stage \(A\alpha\) and  unquote \(\TBL_\alpha M\) is given type
\(\tau\) at stage \(A\alpha\) if \(M\) is given type \(\TW_\alpha \tau\) at
stage \(A\alpha\), represented by the following rules.
\begin{center}
	\infrule{\G\vdash M:\tau @{A\alpha}}{\G\vdash \TB_{\alpha}M:\TW_{\alpha}\tau @A}{\TTB} \andalso
	\infrule{\G\vdash M:\TW_{\alpha}\tau @A}{\G\vdash \TBL_{\alpha}M:\tau @{A\alpha}}{\TTBL}
\end{center}
% \TTB, corresponding to brackets, means 
% if $M$ is typed $\tau$ at stage $A\alpha$ then $\TB_{\alpha}M$, quoted $M$, is typed $\TW_{\alpha}\tau$ at $A$.
% \TTBL\ is converse of \TTB.

@@

\AI{We should discuss conceptual points, not technical points!}

Next, we develop \LMD by extending \LTP with \LLF-like dependent types.
There are three technical points in the extension from \LTP to \LMD.

% Constants and Base Types

First, the way of handling of constants and type-level constants is the difference between \LMD and \LTP.
We adopt a signature $\Sigma$ to handle constants and type-level constants.
This is because a signature simplifies kinding rules relating to type-level constants.
A signature $\Sigma$ is composed of pairs of a base type and its kind or a constant and its type.
For example, if you want to use boolean, $\Sigma = \B::*, \text{true}:\B, \text{false}:\B$

% Kidinding and Well-formed Kinding Rules

Second, we need kinding rules and well-formed kinding rules in order to extend \LMD with dependent types.
It was lucky that almost all rules are determined easily.
This is because multi-stage calculus and dependent types are almost orthogonal.
Therefore, we can get kinding rules and well-formed kinding rules of \LMD just by 
attaching stage anotations to ones of \LLF.
For example, \KAbs-LF is a kinding rule for a dependent type in \LLF and \KAbs\ is a corresponding one.
\begin{center}
	\infrule{\G\vdash \tau :: * \andalso \G,x:\tau @A\vdash \sigma::J}{\G\vdash(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J)}{\KAbs-LF} \\[2mm]
	\infrule{\G\V \tau :: *@A \andalso \G,x:\tau @A\V \sigma::J@A}{\G\V(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J)@A}{\KAbs} \\[2mm]
\end{center}

% Equivalence Rules

Third, we also need type equivalence rules in \LMD because its type system contains dependent types.
Although there are new primitives relating to stages which aren't in \LLF,
we can design all rules easily from the syntax or the reductions of \LMD except one equivalence rule about CSP..

>>>>

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{stage
  variable}\footnote{%
  In Hanada and Igarash~\cite{Hanada2014}, it was called a
  \textit{transition variable}, which is derived from correspondence
  to modal logic, studied by Tsukada and Igarashi~\cite{Tsukada}.} and
a sequence of stage variables is called \textit{stage}.  Intuitively,
they represent the depth of nested brackets.  Stage variables can be
abstracted by $\Lambda\alpha.M$ and instantiated by an application
$M\ A$ to stages.  For example,
$\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10) 5)$ is a code value,
where \(\alpha\) is abstracted.  If it is applied to \(\varepsilon\),
which is the empty sequence standing for zero, \(\TB_\alpha\) becomes
\(\TB_\varepsilon\), which stands for zero brackets, and so, in fact,
\(\TB_\alpha\) disappears.  As a result, the application reduces to
(unquoted) \((\lambda x:\I.x+10) 5\) and to 15.  So, application of a
\(\Lambda\)-abstraction to $\varepsilon$ corresponds to \texttt{run}.
If \(\alpha\) is instantiated with, say, \(\beta\gamma\), then
\(\TB_\alpha\) becomes nested brackets \(\TB_{\beta} \TB_\gamma\).
These are expressed by the following reduction rule:
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
where stage substitution \([\alpha \mapsto A]\) manipulates nesting of
\(\TB\) and \(\TBL\) (and also \(\%_\alpha\) as we see later).

Cross-stage persistence (CSP), which is an important feature of \LTP,
is a primitive to embed values (not necessarily code values) into a
code value.  For example, a \LTP-term
\[
  M_1 = \lambda x:\I.\Lambda\alpha.(\TB_\alpha ((\%_\alpha x) * 2))
\]
takes an integer \(x\) as an input and returns the code value into
which \(x\) is embedded.  CSP is practically important because
one can call library functions from inside quotations.  If the above
\(\lambda\)-abstraction is applied to $1 + 2$ as in
\[
  M_2 = (\lambda x:\I.\Lambda\alpha.(\TB_\alpha ((\%_\alpha x) * 2)))\ (38 + 4)
\]
it evaluates to
\(M_3 = \Lambda\alpha.(\TB_\alpha ((\%_\alpha 42) * 2))\).  According
to the semantics of \LTP, the subterm $\%_\alpha 42$ means that it
waits for the surrounding code to be run (by an application to
$\varepsilon$) and so it does not reduce further.  If \(M_3\) is run
by application to \(\varepsilon\), substitution of \(\varepsilon\) for
\(\alpha\) eliminates \(\TB_\alpha\) and \(\%_\alpha\) and so
\(42 * 2\) is obtained.

The type system of \LTP uses code types---the type of code of type
\(\tau\) is written \(\TW_\alpha \tau\)---for typing \(\TB_\alpha\),
\(\TBL_\alpha\) and \(\%_\alpha\).  It takes stages into account: A
variable declaration in a type environment is associated with its
declared stage as well as its type and a type judgement of \LTP is of
the form $\G \vdash M : \tau @ A$, in which $A$ stands for the stage
of term $M$.  For example,
$y:\I@\alpha \vdash (\lambda x:\I.y) : \I \to \I @ \alpha$ holds, but
$y:\I@\alpha \vdash (\lambda x:\I.y) : \I \to \I @ \varepsilon$ does
not because the latter uses $y$ at stage \(\varepsilon\) but $y$ is
declared at $\alpha$.  Quotation \(\TB_\alpha M\) is given type
\(\TW_\alpha \tau\) at stage \(A\) if \(M\) is given type \(\tau\) at
stage \(A\alpha\); unquote \(\TBL_\alpha M\) is given type \(\tau\)
at stage \(A\alpha\) if \(M\) is given type \(\TW_\alpha \tau\) at
stage \(A\alpha\); and CSP \(\%_\alpha M\) is give type \(\tau\)
at stage \(A\alpha\) if \(M\) is given type \(\tau\) at \(A\).
These are expressed by the following inference rules:
\begin{center}
	\infrule{\G\vdash M:\tau @{A\alpha}}{\G\vdash \TB_{\alpha}M:\TW_{\alpha}\tau @A} \hfil
	\infrule{\G\vdash M:\TW_{\alpha}\tau @A}{\G\vdash \TBL_{\alpha}M:\tau @{A\alpha}} \hfil
	\infrule{\G\vdash M: \tau @A}{\G\vdash \%_{\alpha}M:\tau @{A\alpha}}
\end{center}
% \TTB, corresponding to brackets, means 
% if $M$ is typed $\tau$ at stage $A\alpha$ then $\TB_{\alpha}M$, quoted $M$, is typed $\TW_{\alpha}\tau$ at $A$.
% \TTBL\ is converse of \TTB.

@@

In this paper, we add a simple form of dependent types---{\`a} la
Edinburgh LF~\cite{} and \LLF~\cite{}---to \LTP and develop \LMD.
Types can be indexed by terms as in \texttt{Vector} in
Section~\ref{sec:intro} and \(\lambda\)-abstractions can be given
dependent function types of the form \(\Pi x:\tau. \sigma\) but we do
not consider type operators (such as $\texttt{list } \tau$) or
abstraction over type variables.  We introduce kinds to classify
well-formed types and equivalences for kinds, types, and terms---as
in other dependent type systems---but we have to address a question
how the notion of stage (should) interact with kinds and types.

On the one hand, base types such as \(\I\) should be able to be used
at every stage as in \LTP so that
\(\lambda x:\I.\Lambda \alpha. \TB_\alpha \lambda y:\I.M\) is a valid
term (here, \(\I\) is used at \(\varepsilon\) and \(\alpha\)).
Similarly for indexed types such as Vector 4.  On the other hand, it
is not clear how a type indexed by a variable, which can be used only
at a declared stage, can be used.  For example,
consider
\[\TB_\alpha (\lambda x:\I. (\TBL_\alpha (\lambda y:\text{Vector
  }x.M)N) )
  \text{ and }
  \lambda x:\I. \TB_\alpha (\lambda y:\text{Vector }x.M).
\]
Is Vector\ \(x\) a legitimate type at \(\varepsilon\) (or \(\alpha\),
resp.)  even if \(x:\I\) is declared at stage \(\alpha\) (or
\(\varepsilon\), resp.)?  We give our answer to this question in two
steps.

First, type-level constants such as \(\I\) and Vector can be used at
every stage in \LMD.  Technically, we introduce a signature that
declares kinds of type-level constants and types of constants.  For
example, a signature for the Boolean type and constans is given as
follows $\B::*, \text{true}:\B, \text{false}:\B$ (where $*$ is the
kind of proper types).  Declarations in a siganature are not
associated to particular stages and so they are used at every stage.

Second, an indexed type such as Vector\ 3 or Vector\ $x$ is well
formed only at the stage(s) where the index term is well typed.  Since
constant \(3\) is well typed at every stage (if it is declared in the
signature), so Vector\ 3 is a well-formed type at every stage, too.
However, Vector\ $M$ is well formed only at the stage where index term
$M$ is typed.  Thus, the kinding judgment
of \LMD takes the form \(\G\V \tau :: K @ A\), where stage $A$ stands for
where \(\tau\) is well formed.  For example,
given \(\text{Vector}:: \I \rightarrow *\) in the signature \(\Sigma\),
\(x:\I@\varepsilon \V \text{Vector }x :: * @\varepsilon\) can be
derived but
neither
\(x:\I@\alpha \V \text{Vector }x :: * @\varepsilon\)
nor 
\(x:\I@\epsilon \V \text{Vector }x :: * @\alpha\)
can be.

At first, the restriction above sounds too severe, because a term like
\(\lambda x:\I. \TB_\alpha (\lambda y:\text{Vector }x.M) \), which
models a typical code generator which takes the size $x$ and returns
code for vector manipulation specialized to the given size.  It seems
crucial for \(y\) to be given a type indexed by $x$.  We can address
this problem by CSP---In fact,
$\text{Vector }x$ is not well formed at $\alpha$ under $x:\I@\epsilon$
but $\text{Vector }(\%_\alpha x)$ is!  So, we can still write
\(\lambda x:\I. \TB_\alpha (\lambda y:\text{Vector }(\%_\alpha x).M) \)
for the typical code generator.

Our decision that wellformedness of types takes stages of index terms
into account will lead to the introduction of CSP at the type level
and special equivalence rules, as we will see later.
